---
title: "Analysis of Methylation EPIC Array Data from Human Pancreatic Islets"
author: "Nikolay Oskolkov, SciLifeLab, NBIS Long Term Support, nikolay.oskolkov@scilifelab.se"
date: "January 10, 2020"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
subtitle: PI Charlotte Ling, Lund University
abstract: |
  Here we will use the EPIC Methylation Array data from Human Pancreatic Islets from individuals with and without Type 2 Diabetes (T2D) and perform diebetes status linear predictive analysis with PLS-DA model on 110 manually selected individuals with most reliable phenotypic information.  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="/home/nikolay/WABI/C_Ling/Analysis_110ind_Selected/")
```

### The List of 110 Selected Individuals

All the analysis will be performed on the 110 individuals whic were selected based on multiple criteria such as 1) they are either normo-glycimic or hyper-glycemic / T2D individuals, 2) they have information from all the 4 OMICS (methylation, transcriptomics, phenotypes, genotypes), 3) they all fall within the same age category, and a few other minor criteria. Now we will read the list of those 110 individuals and display a few of them:

```{r}
set.seed(1)
selected_ind<-scan("OVERLAPPING_110_SAMPLES_4OMICS.txt", what = "charater")
selected_ind<-paste0("ID",selected_ind)
print(head(selected_ind, 20))
print(tail(selected_ind, 20))
```


### Checking Human Islets Phenotypes

We will start with the analysis of phenotypes available from approximately 300 Human Panctreatic Islets (but concentrate only on the 110 selected individuals) in order to check the relattion between the phenotypes. The phenotype file below was provided by Human Tissue Lab (HTL) from Lund University Diabetes Center (LUDC), it includes phenotypic information from approximately 300 pancreatic islets donors. We start with reading the file and removing non-relevant phenotypes: 

* HTL_donor_ID: Human Tissue Lab Donor ID)
* NICS_donor_ID: Uppsala donor ID
* Birth: redundant, captured by Age variable
* HbA1c_mmol_mol: redundant, captured by HbA1c_perc
* Blood_group: unlikely to be related to Type 2 Diabetes
* Non_diabetic: redundant, captured by T2D variable
* GAD: not related to T2D
* Gestational: appropriate only for females, will be correlated with Sex
* Date_isolated_islets: unlikely to be related to Type 2 Diabetes
* Date_recieved_islets: unlikely to be related to Type 2 Diabetes
* Islet_equivalences: unlikely to be related to Type 2 Diabetes
* Diabetes_treatment: too sparse information

We will also add information about Methylation EPIC Array batches and RNAseq batches (old vs. new chemistry) in order to see what phenotypes the batches correlate with. This is a potential danger if the batches happen to correlate with valuables phenotypes like T2D or HbA1c. As a control of noise level we will introduce a random variable, we will use it to check later how much of phenotypic variation can be explained by chance.

```{r read phenotypes}
htl<-read.csv("htl_donors.csv",header=TRUE,check.names=FALSE)
htl$HTL_donor_ID<-paste0("ID",htl$HTL_donor_ID)
phen_meth_batch<-read.delim("phenotypes_islets.txt",header=TRUE,row.names=1,check.names=FALSE,sep="\t")
htl$Meth_Batch<-phen_meth_batch$batch[match(as.character(htl$HTL_donor_ID),rownames(phen_meth_batch))]
htl$Meth_Batch[is.na(htl$Meth_Batch)]<-21
phen_rnaseq_batch<-read.delim("phen_rnaseq_batch.txt",header=TRUE,row.names=1,check.names=FALSE,sep="\t")
htl$RNA_Batch<-phen_rnaseq_batch$batch[match(as.character(htl$HTL_donor_ID),rownames(phen_rnaseq_batch))]
htl$Random<-log(sample(dim(htl)[1]))

rownames(htl)<-htl$HTL_donor_ID
htl$HTL_donor_ID<-NULL
htl$NICS_donor_ID<-NULL
htl$Birth<-NULL
htl$HbA1c_mmol_mol<-NULL
htl$Blood_group<-NULL
htl$Non_diabetic<-NULL
htl$GAD<-NULL
htl$Gestational<-NULL
htl$Date_isolated_islets<-NULL
htl$Date_recieved_islets<-NULL
htl$Islet_equivalences<-NULL
htl$Diabetes_treatment<-NULL
head(htl)
```

For convenience we will rename HbA1c_perc to HbA1c, Stimulatory_index to SI, Purity_perc to Purity, Days_cultured to DIC. We will also assign donors with HbA1c > 6.5 to diabetics and balance the data set a bit better in this way. 

```{r rename columns}
names(htl)[names(htl)=="HbA1c_perc"]<-"HbA1c"
names(htl)[names(htl)=="Stimulatory_index"]<-"SI"
names(htl)[names(htl)=="Purity_perc"]<-"Purity"
names(htl)[names(htl)=="Days_cultured"]<-"DIC"
htl$T2D<-ifelse( ((is.na(htl$T2D)==FALSE & htl$T2D==1) | (is.na(htl$HbA1c)==FALSE  & htl$HbA1c>6.5)), 1, 0)
head(htl)
```

Now let us assign "numeric" or "factor" status to the variables. For simplicity, we will treat a variable with less than 2 factor levels as "factor" data type, and else as "numeric" data type. DIC and Meth_Batch can also be considered as factors, however in this case we will need to do a multinomial (not a binomial/logistic) regression which is cumbersome, not clear how to extract adjusted R squared info, and most likely will not bring drammatically different results, so we will stick to only binomial/logistic regression for simplicity.

```{r}
for(i in 1:ncol(htl))
{
  if(length(levels(factor(htl[,i]))) > 2)
  {
    htl[,i]<-as.numeric(as.character(htl[,i]))
  }
  else
  {
    htl[,i]<-as.factor(htl[,i])
  }
}
head(htl)
```

Now we are going to select the 110 individuals with overlapping OMICs and keep only those individuals for further downstream analysis:

```{r}
htl<-htl[match(selected_ind,rownames(htl)),]
head(htl)
```

Now let us calculate pair-wise linear regression for the phenotypes and extract the adjusted R squared information which is equaivalent to the fraction of variation in the response variable explained by the predictor variable. In case the response variable is a factor, we will be using logistic regression (Generalized Linear Model, GLM, with family = "binomial") and calculate the adjusted R suared as:

$$R^2_{adj} = 1 - \frac{\rm{Residual Deviance}}{\rm{Null Deviance}}$$

```{r heatmap phenotypes,fig.width=10,fig.height=8}
htl_adj_r_squared<-matrix(NA,ncol=ncol(htl),nrow=ncol(htl))
for(i in 1:ncol(htl))
{
  print(i)
  for(j in 1:ncol(htl))
  {
    if(typeof(htl[,i])=="double" & (typeof(htl[,j])=="double" | typeof(htl[,j])=="integer"))
    {
      model<-suppressWarnings(lm(htl[,i]~htl[,j]))
      htl_adj_r_squared[j,i]<-suppressWarnings(summary(model)$adj.r.squared)
    }
    if(typeof(htl[,i])=="integer" & typeof(htl[,j])=="double")
    {
      model<-suppressWarnings(lm(htl[,j]~htl[,i]))
      htl_adj_r_squared[j,i]<-suppressWarnings(summary(model)$adj.r.squared)
    }
    if(typeof(htl[,i])=="integer" & typeof(htl[,j])=="integer")
    {
      model<-suppressWarnings(glm(htl[,i]~htl[,j],family="binomial"))
      htl_adj_r_squared[j,i]<-1-model$deviance/model$null.deviance
    }
  }
}
htl_adj_r_squared<-as.data.frame(htl_adj_r_squared)
colnames(htl_adj_r_squared)<-colnames(htl)
rownames(htl_adj_r_squared)<-colnames(htl)
htl_adj_r_squared[htl_adj_r_squared<0]<-0
htl_adj_r_squared
```

Now we can plot the heatmap of the adjusted R squared statistics and check how phenotypes and batches are related to each other:

```{r Phen Heatmap,fig.width=10,fig.height=8}
library("pheatmap")
pheatmap(htl_adj_r_squared, display_numbers=TRUE, fontsize=12, main="Human Pancreatic Islets Phenotypes: Adjusted R^2 of Association")
```

We can see that HbA1c and T2D are very strongly connected, they should since the T2D diagnostics relies on blood glucose level measurements. They also weakly cluster together with BMI implying relation between those three phenotypes, which also makes a lot of sense. Despite Age and Gender are believed to be a risk factor for T2D, we do not observe a reliable association between those three phenotypes in our sample. Meth_Batch and RNA_Batch are moderately connected but happily do not seem to influence any of the important phenotypes, perhaps only the Stimulatory Index (SI). There is one more strange/unexpected observations following from the heatmap DIC seem to be correlated with the T2D status.


### Preparing Methylation Array Data

Let us read the Methylation Array data and have a look:

```{r Read Data}
library("data.table")
met<-suppressWarnings(as.data.frame(fread("methylation_filtered_normalized_no_combat_plus_11_samples_updated_2019_09_19.txt")))
rownames(met)<-met$V1
met$V1<-NULL
met<-subset(met,select=selected_ind)
met<-as.data.frame(t(met))
met[1:6,1:6]
dim(met)
```

Now the methylation data is coupled with the T2D status and ready for performing any statistical analysis such as PCA and PLS-DA.


### Checking Batch-Effects

Let us now check how batch-effects influence the Methylation EPIC Array data. The EPIC array was done using `r length(levels(as.factor(htl$Meth_Batch)))` batches. First of all, let us check if we observe clustering by batches on the PCA plot, i.e. if we observe genome-wide batch-effects:

```{r PCA,fig.width=10,fig.height=8}
library("mixOmics")
X<-met
X<-log10(as.matrix(X)+10)

hist(as.matrix(met),breaks=100, main="Histogram of methylation values before log-transform")
hist(X,breaks=100, main="Histogram of methylation values after log-transform")
X[1:6,1:6]
dim(X)

pca.met<-pca(X,ncomp=10,logratio='none',center=TRUE,scale=TRUE)
pca.met
plot(pca.met,ylim=c(0,0.15))
plotIndiv(pca.met,comp=c(1,2),ind.names=TRUE,group=htl$Meth_Batch,ellipse=FALSE,legend=TRUE,title="PCA EPIC: Batch Effect")
plotIndiv(pca.met,comp=c(1,2),ind.names=TRUE,group=htl$Gender,ellipse=FALSE,legend=TRUE,title="PCA EPIC: Sex Effect")
plotIndiv(pca.met,comp=c(1,2),ind.names=TRUE,group=htl$T2D,ellipse=FALSE,legend=TRUE,title="PCA EPIC: T2D Effect")
plotIndiv(pca.met,comp=c(1,2),ind.names=TRUE,group=htl$DIC,ellipse=FALSE,legend=TRUE,title="PCA EPIC: DIC Effect")
```

We do not seem to see obvious clustering by batch, T2D or DIC on genome-wide level. However this does not mean that each individual CpG is not affected by batch-effects. A very obvious clustering comes from Males vs. Females samples that was present previously has disappeared completely due to the log-transform that made the data more normally-distributed and therefore reduced the batch. 

To further quantify genome-wide batch-effetcs let us display how much of variation in each principal component is explained by the batch variables:

```{r heatmap batch effects,fig.width=10,fig.height=8}
pc_adj_r_squared<-matrix(NA,ncol=dim(pca.met$x)[2],nrow=dim(htl)[2])
for(i in 1:dim(pca.met$x)[2])
{
  print(i)
  for(j in 1:dim(htl)[2])
  {
    pc_adj_r_squared[j,i]<-summary(lm(pca.met$x[,i]~htl[,j]))$adj.r.squared
  }
}
pc_adj_r_squared<-as.data.frame(pc_adj_r_squared)
colnames(pc_adj_r_squared)<-colnames(pca.met$x)
rownames(pc_adj_r_squared)<-colnames(htl)
pc_adj_r_squared[pc_adj_r_squared<0]<-0
pc_adj_r_squared

library("pheatmap")
pheatmap(pc_adj_r_squared, display_numbers=TRUE, fontsize=12, cluster_cols=FALSE, main="Human Islets Methylation Array: Adj R^2 of Association between PCs and Phenotypes")
```

We conclude that PC1 is mostly due to Purity of the Human Pancreatic Islets, while PC2 is mostly due to RNAseq batch which does not make sense at the first glance but my interpretation is that it is somehow confounded by the Methylation batch. The Methylation batch contributes almost only to the PC8, which is good. The PC3,4,5,6,7 the Sex as a main contribution. Overall Sexx seems to be the strongest variable associated with almost all PCs but not PC1 and PC2. The phenotype of interest, T2D and HbA1c are included into PC 5 and especially 8. Let us out of curiosity display top Loadings for PC 5,8, we will do a proper feature selection with PLS-DA and LASSO later, but now we can still get a feeling which CpG sites seem to be correlated with diabetes phenotypes:

```{r,fig.width=10,fig.height=8}
plotLoadings(pca.met,comp=5,method='median',contrib='max',ndisplay=20)
plotLoadings(pca.met,comp=8,method='median',contrib='max',ndisplay=20)
```


### PLS-DA Analysis of Methylation EPIC Array

Now we will perform PLS-DA analysis:

```{r}
gc()
library("mixOmics")
Y<-as.factor(as.character(htl$T2D))
summary(Y)
```

We have approximetaly 29% of T2D individuals which is not crazy unbalanced, so there is a hope that the model does not learn only the majority class. Important to mention that a very naive / stupid classifier that predicts Non-T2D for any given individual should thus achieve 71% accuracy of classification. So our PLS-DA model should be better than the naive classifier and outperform the lower threshold of 71%.

```{r}
my_folds=2
my_nrepeat=5
my_progressBar=FALSE
my_cpus=1
```
```{r perf plsda,fig.width=10,fig.height=8}
gc()
rm(pca.met)
gc()
met_plsda.perf<-plsda(X, Y, ncomp=10)

ptm<-proc.time()
perf.plsda<-perf(met_plsda.perf, validation='Mfold', folds=my_folds, progressBar=my_progressBar, nrepeat=my_nrepeat)
proc.time()-ptm

perf.plsda

head(perf.plsda$error.rate)

head(perf.plsda$error.rate.class)

plot(perf.plsda,overlay='dist',sd=TRUE)
```

BER Mahalanobis and max distances seem to reach their minimum at ncomp=4, so we will use 4 PCs for further PLS-DA analysis.

```{r pls-da,fig.width=10,fig.height=8}
gc()
met_plsda<-plsda(X, Y, ncomp=4)
met_plsda$explained_variance
background = background.predict(met_plsda, comp.predicted=2, dist = "mahalanobis.dist") 
plotIndiv(met_plsda, comp=c(1,2), group=Y, ind.names=TRUE, ellipse=FALSE, background=background, legend=TRUE, title="Human Pancreatic Islets: PLS-DA of Methylation EPIC Array")
```

Interestingly, the samples 220 and 248, that are not T2Ds but end up on the T2D side of the decision boundary, seem to have abnormal BMI values so very likely they were non-diagnosed type 2 diabetics.

```{r}
htl[rownames(htl)=="ID220" | rownames(htl)=="ID248",]
```


```{r pls comp,fig.width=10,fig.height=8}
plotLoadings(met_plsda, comp=1, method='median', contrib='max', ndisplay=20)
plotLoadings(met_plsda, comp=2, method='median', contrib='max', ndisplay=20)
#plotVar(met_plsda, comp=c(1,2), var.names=list(rownames(met)), cex=3)
#cim(met_plsda, row.sideColors=color.mixo(Y), margins=c(8,4))
```


### Sparse PLS-DA Analysis of Methylation EPIC Array

Many of the CpG sites included to the PLS-DA analysis are actually non-informative and bring noise to the analysis. Therefore one needs to apply sparse cleaning algorithm to select a subset of genes that best discriminate between diabetics and non diabetics. We will use LASSO for selection of most informative CpG sites. The tuning procedure below is performed on one component at a time and selects an optimal number of genes that provide lowest error rate. Since from the previous analysis we concluded that the minimal error rate is achieved on the first 4 principal components, we will use ncomp=4 in the sPLS-DA analysis. 

```{r tune splsda,fig.width=10,fig.height=8}
gc()
rm(met_plsda.perf)
gc()
list.keepX<-c(1:10,seq(20,100,10))

ptm<-proc.time()
tune.splsda.islets<-tune.splsda(X,Y,ncomp=4,validation='Mfold',folds=my_folds,progressBar=my_progressBar,dist='mahalanobis.dist',test.keepX=list.keepX,nrepeat=my_nrepeat)
proc.time()-ptm

head(tune.splsda.islets$error.rate)
tune.splsda.islets$choice.keepX
plot(tune.splsda.islets,optimal=TRUE,sd=TRUE)
```

```{r plot splsda,fig.width=10,fig.height=8}
gc()
select.keepX<-as.numeric(tune.splsda.islets$choice.keepX)
splsda.islets<-splsda(X, Y, ncomp=4, keepX=select.keepX)
background = background.predict(splsda.islets, comp.predicted=2, dist = "mahalanobis.dist")
plotIndiv(splsda.islets,comp=c(1,2),group=Y,ind.names=TRUE,ellipse=FALSE,background=background,legend=TRUE,title="Human Pancreatic Islets: PLS-DA of Methylation EPIC Array")

plotVar(splsda.islets,comp=c(1,2),var.names=list(colnames(met)),cex=3)
head(sort(abs(splsda.islets$loadings$X[,"comp1"]),decreasing=TRUE),10)
head(sort(abs(splsda.islets$loadings$X[,"comp2"]),decreasing=TRUE),10)
plotLoadings(splsda.islets,comp=1,method='median',contrib='max',ndisplay=20)
plotLoadings(splsda.islets,comp=2,method='median',contrib='max',ndisplay=20)
cim(splsda.islets,row.sideColors=color.mixo(Y),margins=c(8,4))
```


### Predicting T2D Status from sPLS-DA Model

Finally we will randomly split the methylation data set into training (80% of samples) and validation (20% of samples) sets. This is needed to validate predictions of sPLS-DA trained on the training data set. 

We are going to perform again the tuning of the sPLS-DA model on the training data set. From the previous analysis we know that first 3 PCs minimize the mahalanobis.dist balanced error rate. But here for simplicity since we have only a few samples let us use ncomp=2 and re-tune the optimal numbers of genes although they should be quite similar to the ones obtained in the previous section as the training set comprises 80% of the full data set. For this purpose, we will choose again a small step of the gene grid list.keepX. Now we will apply the tuned and trained model to the validation data set and generate predictions of T2D status and compute the accuracy of the prediction.

The accuracy of T2D vs NonT2D classification is very high, much higher than the 72% accuracy of the naive model. Now we will build confidence intervals by splitting the data set into train and test multiple times and running the PLS-DA classifier for every split.

```{r tune splsda train conf interval,fig.width=10,fig.height=8,eval = TRUE}
gc()
N_repeat<-30
rm(tune.splsda.islets)
gc()
comp1_acc<-vector()
comp2_acc<-vector()
for(k in 1:N_repeat)
{
  print(paste0("Working with split No.", k))
  gc()
  set.seed(k+1000000)
  test_samples<-rownames(met)[sample(1:length(rownames(met)),round(length(rownames(met))*0.2))]
  train_samples<-rownames(met)[!rownames(met)%in%test_samples]
  
  X.train<-X[match(train_samples,rownames(X)),]
  X.test<-X[match(test_samples,rownames(X)),]
  Y.train<-as.factor(as.character(htl[match(train_samples,rownames(htl)),]$T2D))
  Y.test<-as.factor(as.character(htl[match(test_samples,rownames(htl)),]$T2D))
  
  list.keepX<-c(1:10,seq(20,100,10))
  tune.splsda.train.met<-tune.splsda(X.train, Y.train, ncomp=4, validation='Mfold', folds=my_folds,
                                     progressBar=my_progressBar, dist='mahalanobis.dist', logratio="none",
                                     test.keepX=list.keepX, nrepeat=my_nrepeat)
  splsda.met.train<-splsda(X.train, Y.train, logratio='none', ncomp=4,
                           keepX=as.numeric(tune.splsda.train.met$choice.keepX))
  
  splsda.met.predict<-predict(splsda.met.train, X.test, dist='mahalanobis.dist')
  
  print(table(splsda.met.predict$class$mahalanobis.dist[,1],Y.test))
  acc1<-round((sum(diag(table(splsda.met.predict$class$mahalanobis.dist[,1],Y.test)))/sum(table(splsda.met.predict$class$mahalanobis.dist[,1],Y.test)))*100)
  print(paste0("Classification Accuracy from PLS Component 1: ", acc1))
  print(table(splsda.met.predict$class$mahalanobis.dist[,2],Y.test))
  acc2<-round((sum(diag(table(splsda.met.predict$class$mahalanobis.dist[,2],Y.test)))/sum(table(splsda.met.predict$class$mahalanobis.dist[,2],Y.test)))*100)
  print(paste0("Classification Accuracy from PLS Component 2: ", acc2))
  
  comp1_acc<-append(comp1_acc,acc1)
  comp2_acc<-append(comp2_acc,acc2)
  print("***********************************************************")
}
write.table(comp1_acc,file="Comp1_PLS_Meth_Acc.txt",col.names=FALSE,row.names=FALSE,quote=FALSE,sep="\t")
write.table(comp2_acc,file="Comp2_PLS_Meth_Acc.txt",col.names=FALSE,row.names=FALSE,quote=FALSE,sep="\t")
```

```{r,fig.width=10,fig.height=8}
gc()
comp1_acc_arch<-as.numeric(scan("Comp1_PLS_Meth_Acc.txt",what="character"))
comp2_acc_arch<-as.numeric(scan("Comp2_PLS_Meth_Acc.txt",what="character"))

#comp1_acc_arch<-c(comp1_acc,comp1_acc_arch)
#comp2_acc_arch<-c(comp2_acc,comp2_acc_arch)

hist(comp1_acc_arch,breaks=20,xlab="ACCURACY",main="Accuracy T2D Prediction from Methylation: PLS1",col="darkgreen")
mtext(paste0("Accuracy = ",mean(comp1_acc_arch)," +/- ",2*sd(comp1_acc_arch)))
hist(comp2_acc_arch,breaks=20,xlab="ACCURACY",main="Accuracy T2D Prediction from Methylation: PLS2",col="darkgreen")
mtext(paste0("Accuracy = ",mean(comp2_acc_arch)," +/- ",2*sd(comp2_acc_arch)))
```

We can compute how significantly different is the accuracy of the PLS-DA model compared to the naive 71% accuracy:

```{r}
gc()
sum(comp1_acc_arch<=71)/length(comp1_acc_arch)
sum(comp2_acc_arch<=71)/length(comp2_acc_arch)
rm(list=ls())
gc()
```

Finally here is the details on the system on which this document was compiled:

```{r}
sessionInfo()
```
